---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(lubridate)
library(sf)
```

## Load the covid data

First we'll load in the files from the NY Times Gihub repository.
This data has daily totals for COVID cases and deaths both by state and by county.

```{r}
state_cdf <- read_csv("covid-19-data/us-states.csv")
county_cdf <- read_csv("covid-19-data/us-counties.csv")
```

## State and county maps

We'll  also use the `USAboundaries` package to get maps for states and counties.
```{r, eval=FALSE}
install.packages("USAboundaries")  # see below
state_sf <- USAboundaries::us_states() %>%
  rename(fips = statefp)
county_sf <- USAboundaries::us_counties() %>%
  unite("fips", statefp:countyfp, sep = "")
```
Note, that I'm configuring each dataset so that they each have a `fips` column that you can use to join the data.

If the USA boundaries package doesn't work properly, you can download shapefiles for US states counties from the [census]( https://www.census.gov/cgi-bin/geo/shapefiles/index.php). Select the Layer type to be "Counties (or equivalent)" to get counties.
You can do the same for the states by downloading the layer type "States (or equivalent)".
Put the zip files in this folder and unzip them and make sure the unzipped files are in a folder called `tl_2019_us_county` and `tl_2019_us_state`.

__QUESTION__: How could you get the state map from the county map?

Now we can load the shape files.
```{r}
county_sf <- sf::read_sf("tl_2019_us_county/tl_2019_us_county.shp") %>%
  unite("fips", STATEFP:COUNTYFP, sep = "")
state_sf <- sf::read_sf("tl_2019_us_state/tl_2019_us_state.shp") %>%
  rename(fips = STATEFP)
```

The county data is not quite the same as the `USAboundaries` data but each will work well.


## Pulling in US Census data

Finally, let's pull in some census data. First, you'll need a key which is [easy to get](https://api.census.gov/data/key_signup.html).
Once you have your key, install the `tidycensus` package and register your key.



```{r, eval=FALSE}
install.packages("tidycensus")
tidycensus::census_api_key("REPLACE YOUR KEY HERE")
```

First, we need to find the variable that we're looking for.
Download the variable list for the ACS 5-year from 2018 and view it in RStudio.
```{r}
v18 <- tidycensus::load_variables(2018, "acs5", cache = TRUE)
View(v18)
```

You can search for variables using the search bar in the viewer (upper right of the panel).
The variable for population is B01001_001 so let's download that.

```{r}
state_pop <- tidycensus::get_acs("state", "B01001_001") %>%
  rename(fips = GEOID, population = estimate)
county_pop <- tidycensus::get_acs("county", "B01001_001") %>%
  rename(fips = GEOID, population = estimate)
```

Again, I've renamed `GEOID` as `fips` to ensure easy joins.

Feel free to search for other variables in the `v18` table and download those in the same way.

## Putting it together

Now its your turn.

See what you can find out and try making some maps.
I would encourage you to first look at state data since its smaller and then jump into the county level data.
As a group think of some questions that you might be interested in and try to answer them.

Here's a quick example,
```{r}
state_cdf %>% filter(date == max(date)) %>%
  left_join(state_pop %>% select(fips, population)) %>%
  arrange(desc(cases))
```